
\documentclass[a5]{article}
\usepackage{mathpazo} % math & rm
\linespread{1.05}        % Palatino needs more leading (space between lines)
\usepackage[scaled]{helvet} % ss
\usepackage{courier} % tt
%\normalfont
\usepackage[T1]{fontenc}
\usepackage[top=2.0cm,bottom=2.0cm,inner=2.0cm,outer=2.0cm,a5paper,layout=a5paper]{geometry}

\begin{document}
	\pagenumbering{gobble}
	\begin{center}
		{\large \textbf{Propositions}}
		
		\bigskip
		
		accompanying the dissertation
		
		\bigskip
		
		{\bfseries Genetic regulation of gene expression in brain and blood}
		
		\medskip
		
		by
		
		\medskip
		
		{ \textbf{Niek de Klein}}
	\end{center}
	
	\medskip
	
	\begin{enumerate}
		\item Effect sizes of GWAS variants on local molecular traits is often much higher than on disease phenotypes. Knowing what mitigates these large effects is important for understanding disease. \emph{This thesis}
		\item Modelling which cell type contributes most to eQTLs, measured in bulk tissue, can identify disease relevant cell types. \emph{This thesis}
		\item Heterozygous mutations that lower expression of the gene on the affected allele regularly don’t affect overall expression, as the unaffected allele will compensate with higher expression levels. \emph{This thesis}
		\item Making more complicated models that better follow biology can lower statistical power, but increase accuracy. \emph{This thesis}
		\item Genes that function in the brain have stronger selection pressure against genetic variation than genes that function in blood cells.\emph{This thesis}
		\item Without being able to predict cell count proportions interpreting distal genetic regulation of gene expression using bulk samples is impossible. \emph{This thesis}
		\item Multi tissue diseases require multi tissue datasets to identify all causal genes. \emph{This thesis}
		\item Sometimes you need to reinvent the wheel to understand why it is round. 
		\item It’s still magic even if you know how it’s done. \emph{Terry Pratchett, A Hat Full of Sky}.
		

\end{enumerate}

\end{document}
